Uncovering Ferry Traffic Patterns: A Data-Driven Study of Toronto Island Ticket Sales and Redemptions

### 2024-09-27 outputs/paper/paper.qmd
Plots added to section2.3.1
updated peak day summary table, Redemption and Sales Counts Over Time in days, months and a year

library(tidyverse)
library(opendatatoronto)
library(here)
library(ggplot2)

@@

#### Loading the Cleaned Dataset ####
marriage_licence_packages <- search_packages("Toronto Island Ferry Ticket Counts")
marriage_licence_resources <- marriage_licence_packages %>%
  list_package_resources()

@@

# Read the cleaned data
data <- read_csv(file = here("outputs/data/cleaned_data.csv"), 
                 show_col_types = FALSE)

@@

### Temporal Trends in Redemption and Sales Counts

To gain an initial insight into the dataset, we plotted the trends in ticket redemption and sales counts over time. The line plot shown in @fig-linePlot provides a clear visual representation of these trends, serving as a starting point for deeper analysis of ferry usage patterns.

The dataset used in this analysis includes timestamped records of ticket redemptions and sales for the Toronto Island Ferry system, offering a time series perspective on ferry usage. The plot in @fig-linePlot specifically highlights how these counts fluctuate over time, reflecting the flow of passenger demand. The y-axis for redemption counts is displayed in blue, while the y-axis for sales counts is displayed in green, but inverted to better show their correlation.

```{r}
#| label: fig-linePlot
#| fig-cap: Redemption and Sales Counts Over Time
#| echo: false

@@

# Create the plot
ggplot(data, aes(x = Timestamp)) +
  # Line plot for Redemption Count
  geom_line(aes(y = Redemption.Count), color = "#F7C4C1", linewidth = 1.2) +
  # Line plot for Sales Count (inverted)
  geom_line(aes(y = -Sales.Count), color = "#81B0B2", linewidth = 1.2) +

@@

  # Add plot labels
  labs(
    y = "Redemption Count",  # Y-axis for Redemption Count
    x = "Timestamp"          # X-axis for Timestamp
  ) +

  # Apply minimal theme and customize axis titles
  theme_minimal() +
  theme(
    # Color the right-side axis title (Sales Count) to match the orange line
    axis.title.y.right = element_text(color = "#F7C4C1", size = 12),
    # Color the left-side axis title (Redemption Count) to match the pink line
    axis.title.y.left = element_text(color = "#81B0B2", size = 12),
    # Center the title and increase font size
  )
```

As shown in @fig-linePlot, the x-axis denotes the timestamps of each record, while the y-axis represents the number of tickets processed. The blue line corresponds to the redemption counts, and the green line represents the sales counts, plotted inversely to emphasize the contrast between the two trends. The symmetrical y-axis allows for a clearer comparison between peaks and troughs in both counts over time.

The plot reveals significant variations in ferry usage over time. We observe several peaks in both redemption and sales counts, likely aligning with periods of high activity, such as weekends, holidays, or the peak tourist season during the summer months. These spikes suggest increased demand during favorable weather or event-driven occasions. Conversely, the noticeable dips in the graph may correspond to regular weekdays or off-season times, indicating lower ferry usage during these periods. This information is valuable for understanding the temporal dynamics of ferry operations and can inform better resource management and scheduling.

@@

The scatter plot in @fig-scatterPlot is designed to explore the relationship between ticket redemption and sales counts for the Toronto Island Ferry system. This visualization allows us to understand whether there is a correlation between the number of tickets sold and the number of tickets redeemed during the same time period.
```{r}
#| label: fig-scatterPlot
#| fig-cap: Scatter Plot of Redemption vs Sales Count
#| echo: false
#| warning: false
#| message: false

@@

The box plot shown in @fig-boxPlot provides a summary of the distribution of ticket sales and redemption counts for the Toronto Island Ferry system. This visualization helps to identify the spread, central tendency, and any potential outliers in the data for both categories, offering a comparative view of the sales and redemption behaviors.
```{r}
#| label: fig-boxPlot
#| fig-cap: Box Plot of Sales and Redemption Counts
#| echo: false
#| warning: false
#| message: false

@@

In Section 2.2, the time series analysis of ticket sales and redemptions (Figure @fig-linePlot) revealed clear seasonal patterns, with pronounced peaks during the warmer months. This trend aligns with the historical and cultural significance of the Toronto Islands as a recreational destination, as documented by @varga1987toronto. The increased demand during these periods suggests that the ferry system plays a crucial role in supporting both tourism and local leisure activities. These seasonal fluctuations underscore the need for effective resource management and scheduling to accommodate surges in passenger volume during peak times.

>>>>

library(tidyverse)
library(opendatatoronto)
library(here)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)

@@

suppressMessages({
  suppressWarnings({
    #### Loading the Cleaned Dataset ####
    marriage_licence_packages <- search_packages("Toronto Island Ferry Ticket Counts")
    marriage_licence_resources <- marriage_licence_packages %>%
      list_package_resources()

@@

    # Read the cleaned data
    data <- read_csv(file = here("outputs/data/cleaned_data.csv"), 
                     show_col_types = FALSE)
  })
})

@@

### Temporal Trends in Redemption and Sales Counts {#sec-data-temporal}

To gain an initial insight into the dataset, we plotted the trends in ticket redemption and sales counts over time. The line plot shown in @fig-linePlot1, @fig-linePlot2, @fig-linePlot3 and @fig-linePlot4 provides a clear visual representation of these trends, serving as a starting point for deeper analysis of ferry usage patterns.

The peak day analysis table below highlights the highest ferry ticket sales, often aligning with public holidays or major events, such as Aug 1 2016, and July 1 2018. On September 15, 2020, ferry sales saw a significant drop, with only 5,003 tickets sold. This decline could be due to the impact of COVID-19(@lee2021island).

```{r}

#| label: tbl-peak_day
#| echo: false       # Prevents code from being shown
#| warning: false    # Suppresses warnings
#| message: false    # Suppresses messages
#| error: false      # Suppresses errors
#| results: 'asis'   # Ensures the table result is shown without the code

# Load necessary libraries
library(dplyr)
library(lubridate)
library(knitr)

# Step 1: Check if the Timestamp column is properly parsed
data <- data %>%
  mutate(Timestamp = ymd_hms(Timestamp, quiet = TRUE)) %>%  # Quiet mode for parsing errors
  filter(!is.na(Timestamp))  # Remove rows where Timestamp conversion failed

# Step 2: Group data by Year, find the day with maximum Sales for each year
peak_day_per_year_table <- data %>%
  mutate(Year = year(Timestamp)) %>%  # Extract Year from Timestamp
  group_by(Year, Date = as.Date(Timestamp)) %>%  # Group by Year and Date
  summarise(Total_Sales = sum(Sales.Count, na.rm = TRUE), .groups = 'drop') %>%  # Summarize total sales for each day
  filter(Total_Sales > 0) %>%  # Ensure there are sales
  group_by(Year) %>%
  slice_max(order_by = Total_Sales, n = 1, with_ties = FALSE) %>%  # Find the day with maximum sales for each year
  select(Year, Peak_Day = Date, Total_Sales)  # Select relevant columns

# Step 3: Display the table using knitr::kable
kable(peak_day_per_year_table, col.names = c("Year", "Peak Day", "Total Sales"), 
      caption = "Table: Peak Day for Each Year Based on Total Sales")

The dataset used in this analysis provides timestamped records of ticket redemptions and sales for the Toronto Island Ferry system, allowing for a detailed time series analysis of ferry usage patterns. As illustrated in @fig-linePlot1, ferry usage fluctuates significantly throughout the year, with peak activity consistently observed between July and September. This pattern corresponds to the increased passenger demand during the summer holiday season and major events. In 2020, ticket demand saw a sharp decline, likely due to the impact of the COVID-19 pandemic(@urbanyi2020maritime), but it showed signs of gradual recovery starting in 2021 as restrictions eased and tourism began to rebound in Toronto(@travel_gov_2024). By identifying these peak traffic periods in advance, ferry operators can implement targeted operational strategies, such as scheduling additional ferry trips or introducing crowd control measures to minimize the risk of overcrowding or service disruptions. Identifing these high-demand periods can ensure smoother ferry operations and enhance the overall passenger experience.

```{r}
#| label: fig-linePlot1
#| fig-cap: Redemption and Sales Counts Over Time in days
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
#| 
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lubridate)

# Assuming 'data' has columns: 'Timestamp' and 'Sales.Count'
# Convert Timestamp column to proper datetime format and filter out NA values
data <- data %>%
  mutate(
    Timestamp = ymd_hms(Timestamp),   # Convert to datetime format
    Year = year(Timestamp),           # Extract Year
    Month = month(Timestamp, label = TRUE, abbr = TRUE),  # Extract Month (abbreviated)
    Day = day(Timestamp)              # Extract Day
  ) %>%
  filter(!is.na(Year) & !is.na(Month))  # Remove rows with NA in Year or Month

# Group the data by Year, Month, and Day, and calculate the total sales per day for each year
daily_data <- data %>%
  group_by(Year, Month, Day) %>%
  summarise(
    Total_Sales = sum(Sales.Count, na.rm = TRUE),
    .groups = 'drop'  # Ungroup after summarising
  )

# Find the peak sales day for each year
peak_day_data <- daily_data %>%
  group_by(Year) %>%
  filter(Total_Sales == max(Total_Sales)) %>%
  ungroup()

# Step 4: Define colors for each year
year_colors <- c("2015" = "#d07f7a", "2016" = "#377EB8", "2017" = "#7f9f6f", 
                 "2018" = "#984EA3", "2019" = "#FC9272", "2020" = "#EC7014", 
                 "2021" = "#A65628", "2022" = "#C51B7D", "2023" = "#FFD92F")

# Create the plot
ggplot(peak_day_data, aes(x = Month, y = Total_Sales, group = Year, color = as.factor(Year))) +  
  geom_line(size = 1.2) +   # Line plot for each year’s peak day
  geom_point(size = 3) +    # Add points for peak days
  # Set y-axis label and title
  labs(
    x = "Month",
    y = "Total Sales Count",
    color = "Year"
  ) +

  # Manually specify colors for each year
  scale_color_manual(values = year_colors) +

  # Apply minimal theme and customize axis titles
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for better readability
  )

@fig-linePlot2 below shows the Redemption and Sales Counts in months over years 加一段描述plot2

```{r}
#| label: fig-linePlot2
#| fig-cap: Redemption and Sales Counts Over Time in months
#| warnings: false
#| message: false
#| echo: false
#| fig-pos: "H"  #fix the plot in current place when render
#| fig-width: 8
#| fig-height: 4
#| 
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lubridate)

# Convert Timestamp column to proper datetime format and filter out NA values
data <- data %>%
  mutate(
    Timestamp = ymd_hms(Timestamp),     # Convert to datetime format
    Year = year(Timestamp),             # Extract Year
    Month = month(Timestamp, label = TRUE, abbr = TRUE)  # Extract Month (abbreviated)
  ) %>%
  filter(!is.na(Year) & !is.na(Month))  # Remove rows with NA in Year or Month

# Step 2: Group the data by Year and Month and calculate total sales and redemptions per month for each year
monthly_data <- data %>%
  group_by(Year, Month) %>%
  summarise(
    Sales.Count = sum(Sales.Count, na.rm = TRUE),
    Redemption.Count = sum(Redemption.Count, na.rm = TRUE),
    .groups = 'drop'  # Ungroup after summarising
  )

# Step 3: Determine the maximum value for symmetrical scaling
max_count <- max(max(monthly_data$Sales.Count), max(monthly_data$Redemption.Count))

# Step 4: Define colors for each year
year_colors <- c("2015" = "#d07f7a", "2016" = "#377EB8", "2017" = "#7f9f6f", 
                 "2018" = "#984EA3", "2019" = "#FC9272", "2020" = "#EC7014", 
                 "2021" = "#A65628", "2022" = "#C51B7D", "2023" = "#FFD92F")

# Step 5: Create the plot with inverted Redemption Count for bottom half visualization
ggplot(monthly_data, aes(x = Month)) +  
  # Line plot for Redemption Count (inverted)
  geom_line(aes(y = -Redemption.Count, color = as.factor(Year), group = Year), linewidth = 1.2) +
  # Line plot for Sales Count (normal)
  geom_line(aes(y = Sales.Count, color = as.factor(Year), group = Year), linetype = "dashed", linewidth = 1.2) +

  # Set symmetrical y-axis limits for both Sales and Redemption
  scale_y_continuous(limits = c(-max_count, max_count),
                     breaks = seq(-max_count, max_count, by = 100000),
                     sec.axis = sec_axis(~ -., name = "Sales Count (dashed lines)")) +

  # Add plot labels
  labs(
    y = "Redemption Count (straight lines)",  # Y-axis for Redemption Count (inverted)
    x = "Month",             # X-axis for Month
    color = "Year"
  ) +

  # Manually specify colors for each year
  scale_color_manual(values = year_colors) +

  # Apply minimal theme and customize axis titles
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "black", size = 12, margin = margin(r = 10)),
    axis.title.y.left = element_text(color = "black", size = 12, margin = margin(l = 10)),
    plot.title = element_text(hjust = 0.5, size = 16),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank()
  )

@fig-linePlot3 on the next page compares the sales count and redemption count in year 2021 and 2023

1 emm这里写一点描述fig3的对比？，render后看页面排版

```{r}
#| label: fig-linePlot3
#| fig-cap: Redemption and Sales Counts Over Time in 2021 and 2023
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"  #fix the plot in current place when render
#| fig-width: 8
#| fig-height: 4

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lubridate)
library(gridExtra)  # To arrange plots

# Filter the data for 2021 and 2023
data_2023 <- data %>%
  mutate(Timestamp = ymd_hms(Timestamp)) %>%
  filter(year(Timestamp) == 2023)

data_2021 <- data %>%
  mutate(Timestamp = ymd_hms(Timestamp)) %>%
  filter(year(Timestamp) == 2021)

# Calculate the range for both y-axes for symmetry (using both years)
y_max <- max(max(data_2021$Sales.Count), max(data_2021$Redemption.Count),
             max(data_2023$Sales.Count), max(data_2023$Redemption.Count))

# Plot for 2023
plot_2023 <- ggplot(data_2023, aes(x = Timestamp)) +
  geom_line(aes(y = Redemption.Count), color = "#A977A6", linewidth = 1.2) +
  geom_line(aes(y = -Sales.Count), color = "#C1A9D2", linewidth = 1.2) +
  scale_y_continuous(limits = c(-y_max, y_max), sec.axis = sec_axis(~ -., name = "Sales Count")) +
  scale_x_datetime(date_breaks = "4 months", date_labels = "%b") +  # Show only 3 labels
  labs(y = "Redemption Count", x = "Month", title = "2023") +
  theme_minimal()

# Plot for 2021
plot_2021 <- ggplot(data_2021, aes(x = Timestamp)) +
  geom_line(aes(y = Redemption.Count), color = "#F49600", linewidth = 1.2) +
  geom_line(aes(y = -Sales.Count), color = "#F9CB80", linewidth = 1.2) +
  scale_y_continuous(limits = c(-y_max, y_max), sec.axis = sec_axis(~ -., name = "Sales Count")) +
  scale_x_datetime(date_breaks = "4 months", date_labels = "%b") +  # Show only 3 labels
  labs(y = "Redemption Count", x = "Month", title = "2021") +
  theme_minimal()

# Arrange the two plots side by side
grid.arrange(plot_2021, plot_2023, ncol = 2)

```
As shown in @fig-linePlot4, the x-axis denotes the timestamps of each record, while the y-axis represents the number of tickets processed. The blue line corresponds to the redemption counts, and the green line represents the sales counts, plotted inversely to emphasize the contrast between the two trends. The symmetrical y-axis allows for a clearer comparison between peaks and troughs in both counts over time.

```{r}
#| echo: false 
#| warning: false
#| message: false
#| label: fig-linePlot4
#| fig-cap: Redemption and Sales Counts Over Time in years
#| fig-pos: "H"  #fix the plot in current place when render
#| fig-width: 6
#| fig-height: 2.5

@@

# Create the plot
ggplot(data, aes(x = Timestamp)) +
  # Line plot for Redemption Count
  geom_line(aes(y = Redemption.Count), color = "#F7C4C1", linewidth = 1.2) +
  # Line plot for Sales Count (inverted)
  geom_line(aes(y = -Sales.Count), color = "#83A0BE", linewidth = 1.2) +

@@

  # Add plot labels
  labs(
    y = "Redemption Count",  # Y-axis for Redemption Count
    x = "Years"          # X-axis for Timestamp
  ) +

  # Apply minimal theme and customize axis titles
  theme_minimal() +
  theme(
    # Color the right-side axis title (Sales Count) to match the orange line
    axis.title.y.right = element_text(color = "#F7C4C1", size = 12),
    # Color the left-side axis title (Redemption Count) to match the pink line
    axis.title.y.left = element_text(color = "#83A0BE", size = 12),
    # Center the title and increase font size
  )

The plot reveals significant variations in ferry usage over years. We observe several peaks in both redemption and sales counts, likely aligning with periods of high activity, such as weekends, holidays, or the peak tourist season during the summer months. These spikes suggest increased demand during favorable weather or event-driven occasions. Conversely, the noticeable dips in the graph may correspond to regular weekdays or off-season times, indicating lower ferry usage during these periods. This information is valuable for understanding the temporal dynamics of ferry operations and can inform better resource management and scheduling.

@@

The scatter plot in @fig-scatterPlot is designed to explore the relationship between ticket redemption and sales counts for the Toronto Island Ferry system. This visualization allows us to understand whether there is a correlation between the number of tickets sold and the number of tickets redeemed during the same time period.

```{r}
#| label: fig-scatterPlot
#| fig-cap: Scatter Plot of Redemption vs Sales Count
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 3

@@

The box plot shown in @fig-boxPlot provides a summary of the distribution of ticket sales and redemption counts for the Toronto Island Ferry system. This visualization helps to identify the spread, central tendency, and any potential outliers in the data for both categories, offering a comparative view of the sales and redemption behaviors.
```{r}
#| label: fig-boxPlot
#| fig-cap: Box Plot of Sales and Redemption Counts
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 2.5

@@

In @sec-data-temporal, the time series analysis of ticket sales and redemptions (Figure @fig-linePlot1, @fig-linePlot2, @fig-linePlot3 and  @fig-linePlot4 ) revealed clear seasonal patterns, with pronounced peaks during the warmer months. This trend aligns with the historical and cultural significance of the Toronto Islands as a recreational destination, as documented by @varga1987toronto. The increased demand during these periods suggests that the ferry system plays a crucial role in supporting both tourism and local leisure activities. These seasonal fluctuations underscore the need for effective resource management and scheduling to accommodate surges in passenger volume during peak times.
